---
title: "STAT 886 HW 1"
author: "David Nguyen"
date: "February 8, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# 1.3

Probability calculation for genetics (from Lindley, 1965): suppose that in each individual of a large population there is a pair of genes, each of which can be either x or X, that controls eye color: those with xx have blue eyes, while heterozygotes (those with Xx or xX) and those with XX have brown eyes. 
The proportion of blue-eyed individuals is p^2^ and of heterozygotes is 2p(1 − p), where 0 < p < 1. Each parent transmits one of its own genes to the child; if a parent is a heterozygote, the probability that it transmits the gene of type X is 1/2 . 

**Assuming random mating, show that among brown-eyed children of brown-eyed parents, the expected proportion of heterozygotes is 2p/(1+2p).** 

Probabilities of heterozygote child given both parent's have brown eyes

* p(xX | XX, XX) = 0
* p(xX | xX, XX) = 1/2
* p(xX | xX, xX) = 1/2

Prior probabilities of parent chromosomal combinations

* p(XX, XX) = (1-p)^2^(1-p)^2^ = (1-p)^4^
* p(xX, XX) = (1-p)^2^2p(1-p) + 2p(1-p)(1-p)^2^  = 4p(1-p)^3^
* p(xX, xX) = 2p(1-p)2p(1-p) = 4p^2^(1-p)^2^

probability a child has brown eyes given their parent's

* p(brown | XX,XX) = 1
* p(brown | xX, XX) = 1
* p(brown | xX, xX) = 3/4

\begin{align}
p(\text{xX, child has brown eyes, parents have brown eyes}) & = \frac{p(\text{xX, parents have brown eyes})}{p(\text{child has brown eyes, parents have brown eyes})} \\
                                      & = \frac{0 \cdot (1-p)^4 + 1/2 \cdot 4p(1-p)^3 + 1/2 \cdot 4p^2(1-p)^2}{1 \cdot (1-p)^4 + 1 \cdot 4p(1-p)^3 + 3/4 \cdot 4p^2(1-p)^2} \\
                                      & = \frac{2p(1-p)^3 + 2p^2(1-p)^2}{(1-p)^4 + 4p(1-p)^3 + 3p^2(1-p)^2} \\
                                      & = \frac{2p(1-P) + 2p^2}{(1-p)^2 + 4p(1-p)+xp^2} \\
                                      & = \frac{2p - 2p^2 + 2p}{p^2 - 2p + 1 + 4p - 4p^2 + 3p^2} \\
                                      & = \frac{2p}{1+2p}
\end{align}

**Suppose Judy, a brown-eyed child of brown-eyed parents, marries a heterozygote, and they have
*n* children, all brown-eyed. Find the posterior probability that Judy is a heterozygote.**

From the previous problem we know $P(\text{Judy is xX} | \text{her and her parents eyes are brown} = \frac{2p}{1+2p}$ which implies that $P(\text{Judy is XX} | \text{her and her parents eyes are brown} = \frac{1}{1+2p}$, The probability that Judy and her heterozygous partner produce $n$ children with brown eyes is $\left(\frac{3}{4}\right)^n$ given that she is a heterozygote and $1^n$ given that she is a homozygote (XX).

\begin{align}
P(\text{Judy is xX} | \text{Judy, her parents, and n children have brown eyes}) & =
                    \frac{\frac{2p}{1+2p} \left(\frac{3}{4}\right)^n}{\frac{2p}{1+2p} \left(\frac{3}{4}\right)^n + \frac{1}{1+2p} 1^n}
\end{align}

# 1.4
Probability assignment: we will use the football dataset to estimate some conditional
probabilities about professional football games. There were twelve games with point
spreads of 8 points; the outcomes in those games were: −7, −5, −3, −3, 1, 6, 7, 13, 15,
16, 20, and 21, with positive values indicating wins by the favorite and negative values
indicating wins by the underdog. Consider the following conditional probabilities:

* Pr(favorite wins | point spread = 8),
* Pr(favorite wins by at least 8 | point spread = 8),
* Pr(favorite wins by at least 8 | point spread = 8 and favorite wins).

### 1.4a
Estimate each of these using the relative frequencies of games with a point spread of
8.

\begin{align}
Pr(\text{favorite wins | point spread = 8}) &= \frac{8}{12} \\
Pr(\text{favorite wins by at least 8 | point spread = 8}) & = \frac{5}{12} \\
Pr(\text{favorite wins by at least 8 | point spread = 8 and favorite wins}) &= \frac{5}{8}
\end{align}

### 1.4b
Estimate each using the normal approximation for the distribution of (outcome − point spread).

```{r}
scorediff <- c( −7, −5, −3, −3, 1, 6, 7, 13, 15, 16, 20, 21)
d <- scorediff - 8
sdd <- round(sd(d))
```

The sample standard deviation of deviations (score difference - point spread) is `r round(sd(d),2)` which we will round to `r sdd`. Even though the sample mean of deviations is `r mean(d)` we will assume, following BDA3, that a mean of zero is appropriate for approximating the deviations such that $d \sim N(0, 10^2)$.

```{r}
prob1 <- 1 - pnorm(-8, mean = 0, sd = sdd)
prob2 <- 1 - pnorm(0, mean = 0, sd = sdd)
prob3 <- prob2 / prob1
```

* Pr(favorite wins | point spread = 8) = `r prob1`
* Pr(favorite wins by at least 8 | point spread = 8) = `r prob2`
* Pr(favorite wins by at least 8 | point spread = 8 and favorite wins) = `r prob3`

# 1.6
Conditional probability: approximately 1/125 of all births are fraternal twins and 1/300
of births are identical twins. Elvis Presley had a twin brother (who died at birth). What
is the probability that Elvis was an identical twin? (You may approximate the probability
of a boy or girl birth as 1/2.)

Note that:

* pr(identical twin brothers) = (1/300)(1/2) = 1/600
* pr(fraternal twin brothers) = (1/125)(1/2)^2^ = 1/500
* pr(twin brothers) = 1/600 + 1/500 = 11/3000

So,

\begin{align}
pr(\text{Elvis' brother was identical}) & = \frac{pr(\text{identical twin brothers})}{pr(\text{twin brothers})} \\
                                        & = \frac{1/600}{11/3000} \\
                                        & = \frac{5}{11}
\end{align}

# 1.7
Conditional probability: the following problem is loosely based on the television game
show Let’s Make a Deal. At the end of the show, a contestant is asked to choose one of
three large boxes, where one box contains a fabulous prize and the other two boxes contain
lesser prizes. After the contestant chooses a box, Monty Hall, the host of the show,
opens one of the two boxes containing smaller prizes. (In order to keep the conclusion
suspenseful, Monty does not open the box selected by the contestant.) Monty offers the
contestant the opportunity to switch from the chosen box to the remaining unopened box.
Should the contestant switch or stay with the original choice? Calculate the probability
that the contestant wins under each strategy. This is an exercise in being clear about the
information that should be conditioned on when constructing a probability judgment.
See Selvin (1975) and Morgan et al. (1991) for further discussion of this problem.



